---
title: "SAE - Unit Level Model"
subtitle: "Modeling Poverty Rates at the Unit Level"
author: "Mamadou S Diallo, Ph.D"
date: today
format:
  html:
    theme: readable
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---


```{r}

library(dplyr)
library(emdi)
library(gt)
library(haven)
library(readxl)
library(srvyr)
library(tibble)

```


## FGT Poverty Measures 

The Foster-Greer-Thorbecke (FGT) class of poverty measures is a widely used set of indicators in poverty analysis. The FGT measures are defined as follows:

$$ F_{\alpha d} = \frac{1}{N_d} \sum_{j=1}^{N_d} \left( \frac{z - E_{dj}}{z} \right)^\alpha I(E_{di} < z) $$

where:

- $N_d$ is the total population in area $d$, 
- $E_{dj}$ is the income or consumption expenditure of individual or household $j$ in area $d$, 
- $z$ is the poverty line, $I(E_{di} < z)$ is an indicator function that equals 1 if the income or expenditure of individual or household $j$ is below the poverty line and 0 otherwise, and 
- $\alpha$ is a parameter that determines the sensitivity of the measure to poverty depth.
     - $\alpha = 0$ : Headcount Ratio (HCR) - measures the proportion of the population below the poverty line.
     - $\alpha = 1$ : Poverty Gap Index (PGI) - measures the average  
    proportionate shortfall of the total population from the poverty line.
     - $\alpha = 2$ : Squared Poverty Gap Index (SPGI) - gives more weight to those who are further below the poverty line, emphasizing the severity of poverty.

It is assumed that there exists a one-to-one transformation $Y_{dj} = T(E_{dj})$ of the welfare variables, $E_{dj}$ , such that the vector $Y_d = (y_{d1}, ..., y_{dN_d})^T ∼ N(\mu_d
, \Sigma_d)$

For example, we may have $T(x) = log(x)$.

## Wrangling and Combining the Datasets

```{r}

pov_df <- 
  haven::read_sav("../data/eicvmb2020/BASE PAUVRETE.sav") |>
  dplyr::rename_with(tolower) |>
  dplyr::arrange(id01, id02)
  

communes_df <- 
  readxl::read_excel("../data/communes_data.xlsx", sheet="menages") |>
  dplyr::rename_with(tolower) |>
  dplyr::arrange(id_commune, id_menage)

# commune_names_df = readxl::read_excel("../data/communes_data.xlsx", sheet="communes")

```

### Inspecting the Datasets

Definition of the communes

```{r}

dim(communes_df)
head(communes_df)

```

Poverty Survey Data

```{r}

dim(pov_df)

pov_df |>
  dplyr::select(idm, id_zd, id01, id02, id03) |>
  head(7)

```

Merging the datasets to add the communes to the survey data

```{r}

pov_merged_df <- 
  pov_df |>
  dplyr::left_join(
    communes_df,
    by = c("idm" = "id_menage")
  ) |>
  dplyr::rename(urban = milieu)


pov_merged_df |> 
  dplyr::select(idm, id01, id_commune, nom_commune) |>
  head(7)

```


## Auxiliary Variables

As seen before, we need to identify auxiliary variables at the unit (household) area level that are:      

 - **correlated with the target variable** (poverty status),     
 - **available from census or administrative data or Big data**.


For illustration, we will consider these auxiliary variables available from census data at the commune level:

 - `urban` : Urban / Rural indicator
 - `hhsize` : household size
 - `n_rooms` : number of rooms


```{r}

hh_census <- 
  haven::read_sav("../data/rgph2008/habit.sav") |>
  dplyr::rename_with(tolower) |>
  dplyr::select(
    commune, urhabit, population, h02
  ) |>
  dplyr::mutate(
    commune = as.integer(commune)
  )

head(hh_census) |> gt::gt()

```

Creating the indicators

```{r}

hh_census <- 
  hh_census |>
  dplyr::mutate(
    urban = ifelse(urhabit == 1, 1, 0)
  ) |>
  dplyr::rename(
    hhsize = population,
    n_rooms = h02
  )

head(hh_census) |> gt::gt()

```

## Removing Missing Communes

**Issue**: Some communes in the sample are not in the census. Maybe, it is a nomenclature issue.

```{r}

communes_not_in_census <- 
  pov_merged_df |>
  dplyr::filter(
    !(id_commune %in% hh_census$commune)
  )

head(communes_not_in_census) |> gt::gt()

```

Let's exclude these communes for now

What are the common communes between the two datasets?

```{r}

intersect_communes <- 
  intersect(
    pov_merged_df$id_commune,
    hh_census$commune
  )

```

Let's filter both datasets to keep only these communes and then combine them

```{r}

pov_merged_df2 <- 
  pov_merged_df |>
  dplyr::filter(
    id_commune %in% intersect_communes
  )

hh_census2 <- 
  hh_census |>
  dplyr::filter(
    commune %in% intersect_communes
  )

```


## Unit Level Model Estimation


```{r}

umodel_pov_abs <- 
  emdi::ebp(
    fixed = tot_agg_def ~ as.factor(urban) + hhsize + n_rooms,
    pop_data = hh_census2,
    pop_domains = "commune", 
    smp_data = pov_merged_df2,
    smp_domains = "id_commune",
    threshold = 11091.37, # poverty line in 2020
    transformation = "log",
    weights = "weight",
    #interval = c(1100,310500),
    L = 15, # much larger number in practice
    MSE = TRUE,
    B = 50, # much larger number in practice
    seed = 123,
    na.rm = TRUE,
    # custom_indicator = list(
    #   HCR_singleHH = function(y, threshold) {
    #                           mean(y < threshold)
    #                     }
    # )
  )

summary(umodel_pov_abs)

```




## Did the CVs Improve ?

```{r}

est <- umodel_pov_abs$ind |>
  dplyr::select(Domain, Head_Count) |>
  dplyr::rename(
    est_taux = Head_Count
  )

mse <- umodel_pov_abs$MSE |>
  dplyr::select(Domain, Head_Count) |>
  dplyr::rename(
    mse_taux = Head_Count
  )
  
cvs <- est |>
  dplyr::inner_join(
    mse,
    by = "Domain"
  ) |>
  dplyr::mutate(
    cv_model = sqrt(mse_taux) / (est_taux + 1e-10),
  ) |>  
  dplyr::arrange(desc(cv_model))


cvs |>
  gt::gt() |>
  gt::fmt_percent(
    columns = c(cv_model),
    decimals = 1
  ) |>
  gt::cols_label(
    Domain = "Commune",
    est_taux = "Model Estimate",
    mse_taux = "Model MSE",
    cv_model = "Model CV"
  )

```



